---
title: "Entering many records at once"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Entering many records at once}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A [shinyformtools](https://github.com/Gwstat/shinyformtools) form describes one
record at a time: a dialog opens, the user fills it in, one row is stored.
Some data does not come in single records. A bird survey visit produces a
dozen numbers at once, one per species and distance band; sending survey
teams out means handing out equipment from a shared store. shinygridtools
adds two tools for that, built on the same form description:

1. **the grid**, which shows a group of records as a spreadsheet and saves
   what the user changed;
2. **the basket**, a field whose items come from a stock that every record
   shares.

This vignette walks through both and shows what they store. The database
parts below run as written, against a throwaway SQLite file; the app parts
are shown as code.

```{r library, message = FALSE}
library(shinygridtools)  # attaches shinyformtools as well
```

## 1. The records behind a grid

A grid needs nothing beyond an ordinary form. Here every stored record is one
species at one site on one visit, with two counts:

```{r form}
db <- db_sqlite(tempfile(fileext = ".sqlite"))

counts <- form(
  form_id = "counts",
  table_name = "counts",
  db = db,
  fields = list(
    form_field(id = "visit", label = "Visit"),
    form_field(id = "site", label = "Site"),
    form_field(id = "species", label = "Species"),
    form_field(id = "adults", label = "Adults", input_type = "numericInput"),
    form_field(id = "juveniles", label = "Juveniles", input_type = "numericInput")
  )
)

conn <- db_connect(db)
init_db(counts, conn = conn)
```

`grid_server()` takes three things from it: the **group** that one grid shows
(`visit` and `site`), the **row key** (`species`, one grid row per species)
and the **columns** (by default every numeric field that is neither group nor
key, here `adults` and `juveniles`).

## 2. The grid in an app

```{r app, eval = FALSE}
ui <- fluidPage(
  selectInput("site", "Site", c("Pond", "Meadow")),
  grid_ui("counts", label = "Birds per species")
)

server <- function(input, output, session) {
  grid_server(
    "counts", counts,
    rows = c("Blackbird", "Robin", "Great tit"),
    key = "species",
    group = reactive(list(visit = "1", site = input$site)),
    user = "demo"
  )
}

shinyApp(ui, server)
```

Enter and the arrow keys walk the cells, a block copied from a spreadsheet is
pasted from the focused cell onwards, and sums follow the entries. Half a
second after the last keystroke the grid saves. Choosing another site shows
that site's records.

## 3. What a save does

A save writes only what the user changed, through
`shinyformtools::upsert_records()`, in one transaction. A row that gets values
is inserted, a changed row updated, a row the user emptied soft-deleted. The
grid calls it for you; this is what it hands over when someone fills in two
species at the pond:

```{r save}
upsert_records(
  counts,
  list(
    list(visit = "1", site = "Pond", species = "Blackbird", adults = 4, juveniles = 10),
    list(visit = "1", site = "Pond", species = "Robin", adults = 2)
  ),
  key = c("visit", "site", "species"),
  conn = conn, user = "ada"
)

fetch_records(counts, conn = conn)[, c("site", "species", "adults", "juveniles")]
```

Each write is an ordinary record write: validated against the form's rules,
audit-logged, visible in a `form_server()` table of the same form.

### Two people, one site

The grid remembers what it showed and sends it along as `expected`. When
someone else changed a cell in the meantime, that record is not overwritten:

```{r conflict}
# Bob corrects the blackbirds while Ada still sees 4 adults.
invisible(update_record(counts, list(adults = 5), record_id = 1, conn = conn, user = "bob"))

result <- tryCatch(
  upsert_records(
    counts,
    list(list(visit = "1", site = "Pond", species = "Blackbird", adults = 6)),
    key = c("visit", "site", "species"),
    expected = list(list(adults = 4)),
    conn = conn, user = "ada"
  ),
  sft_edit_conflict = function(e) conditionMessage(e)
)
result

fetch_records(counts, conn = conn)$adults
```

In the app, the grid takes such a row out, stores the others, tells the user
which row someone else changed and shows the current values. With `poll` (on
by default, every 3 seconds) it also shows what others saved, cell by cell,
without touching the cell the user is typing in.

## 4. One record per cell

Data is often stored long: one record per species *and* distance band,
holding a single number. `col_key` and `value` switch the grid to that
layout; the columns are then the values of `col_key`:

```{r long, eval = FALSE}
grid_server(
  "bands", observations,
  rows = c("Blackbird", "Robin"), key = "species",
  cols = c("Under 25 m" = "1", "25 to 100 m" = "2", "Over 100 m" = "3"),
  col_key = "band", value = "count",
  group = reactive(list(visit = input$visit, site = input$site))
)
```

An emptied cell soft-deletes its record; two people typing into different
cells of one row never collide.

## 5. A grid inside one record

Sometimes the whole table belongs to one record: the tally sheet of one
visit. `grid_input` is then a field type, and the matrix is stored as one
JSON text with its row and column names:

```{r grid-field}
sheets <- form(
  form_id = "sheets",
  table_name = "sheets",
  db = db,
  fields = list(
    form_field(id = "site", label = "Site"),
    form_field(
      id = "tally", label = "Tally", input_type = "grid_input",
      args = list(rows = c("Blackbird", "Robin"), cols = c("Adults", "Juveniles"))
    )
  )
)
init_db(sheets, conn = conn)

tally <- matrix(
  c(12, 7, 4, 3), nrow = 2,
  dimnames = list(c("Blackbird", "Robin"), c("Adults", "Juveniles"))
)
invisible(insert_record(sheets, list(site = "Pond", tally = tally), conn = conn))

fetch_records(sheets, conn = conn)$tally
```

Stored by name, the grid survives a change of `rows`: reordered rows land in
the right place, and rows the field no longer lists are still shown while they
hold numbers.

## 6. The basket and its stock

A basket field holds items from a palette: dragged or clicked in, counted
with `+` and `-`, marked as kept on site. The palette is another form,
named with `cart_catalog()`:

```{r basket}
equipment <- form(
  form_id = "equipment", table_name = "equipment", db = db,
  fields = list(
    form_field(id = "code", label = "Code", unique = TRUE),
    form_field(id = "name", label = "Name"),
    form_field(id = "stock", label = "Stock", input_type = "numericInput")
  )
)

teams <- form(
  form_id = "teams", table_name = "teams", db = db,
  fields = list(
    form_field(id = "team", label = "Team"),
    form_field(
      id = "kit", label = "Equipment", input_type = "cart_input",
      args = list(catalog = cart_catalog(equipment, id = "code", label = "name", stock = "stock"))
    )
  )
)

init_db(equipment, conn = conn)
init_db(teams, conn = conn)
invisible(insert_record(equipment, list(code = "scope", name = "Spotting scope", stock = 3), conn = conn))
```

The stock is shared by every basket that names the catalog. A save that
would take more than is left is refused, from the dialog, from the board and
from scripts alike:

```{r stock}
invisible(insert_record(teams, list(
  team = "Pond team",
  kit = data.frame(item = "scope", n = 2)
), conn = conn))

tryCatch(
  insert_record(teams, list(
    team = "Meadow team",
    kit = data.frame(item = "scope", n = 2)
  ), conn = conn),
  sft_validation_error = function(e) conditionMessage(e)
)
```

Lines kept on site (`on_site = TRUE`) do not count against the stock, and
deleting a record frees what it held. On MariaDB two people taking the last
piece at the same moment are serialised by a lock per stock, so only one of
them gets it.

## 7. The board

`cart_board_ui()` (or `form_ui(show_board = TRUE)` of shinyformtools) puts
every record next to the stock. One drag moves one piece: from the stock to a
team, between teams, or back. Each move is saved at once, with the same
checks as the dialog.

```{r board, eval = FALSE}
ui <- fluidPage(
  form_ui("teams", title = "Survey teams", show_board = TRUE)
)

server <- function(input, output, session) {
  form_server("teams", teams, user = "demo")
}
```

```{r cleanup, include = FALSE}
db_disconnect(conn)
```

## Where to go next

- `run_example("app_grid_entry")` is a bird survey: counts per site, species
  and distance band on two visits, with the first visit's counts greyed
  inside the cells as a hint.
- `run_example("app_cart_input")` lends equipment to survey teams from a
  shared store, with the board above the table.
- `?grid_module`, `?grid_input` and `?cart_input` document every argument.
- The shinyformtools vignette explains the form description, the database it
  produces and how to read it without the app.
